---
title: "Allegheny Crash Study"
author: "Lauren Renaud"
date: "June 9, 2016"
output: 
  html_document:
    toc: true
    toc_depth: 4
    theme: cerulean
    highlight: tango
---

```{r global_options, include=FALSE}
library(ggplot2)
library(ISLR)
library(MASS)
library(klaR)
library(knitr)
library(boot)
library(leaps)
library(caret)
library(glmnet)
library(plyr)
library(gam)
library(gridExtra)
library(partykit)
library(rpart)
library(rattle)
library(randomForest)
library(pROC)
library(rpart.plot)
library(reshape2)
library(dplyr)

knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
options (scipen=4)
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```


```{r import_data, cache=TRUE}
crash15 <- read.csv("2014alcocrash.csv", header=TRUE, sep = ",")
crash14 <- read.csv("2014alcocrash.csv", header=TRUE, sep = ",")
crash13 <- read.csv("2013alcocrash.csv", header=TRUE, sep = ",")
crash12 <- read.csv("2012alcocrash.csv", header=TRUE, sep = ",")
crash11 <- read.csv("2011alcocrash.csv", header=TRUE, sep = ",")
crash10 <- read.csv("2010alcocrash.csv", header=TRUE, sep = ",")
crash09 <- read.csv("2009alcocrash.csv", header=TRUE, sep = ",")
crash08 <- read.csv("2008alcocrash.csv", header=TRUE, sep = ",")
crash07 <- read.csv("2007alcocrash.csv", header=TRUE, sep = ",")
crash06 <- read.csv("2006alcocrash.csv", header=TRUE, sep = ",")
crash05 <- read.csv("2005alcocrash.csv", header=TRUE, sep = ",")
crash04 <- read.csv("2004alcocrash.csv", header=TRUE, sep = ",")


crashes <- rbind(crash04, crash05, crash06, crash07, crash08, crash09, crash10, crash11, crash12, crash13, crash14, crash15)
# write.table(crashes, "/Users/lauren/Documents/School/CMU/Box Sync/Clubs/SUDS/County Crash Data/crashes.csv", sep=",", row.names = FALSE)

# crashes <- read.csv("/Users/lauren/Documents/School/CMU/Box Sync/Clubs/SUDS/County Crash Data/allto2015alcocrash.csv", sep=",", header=TRUE)

### Features Engineering

# quarter off day hours
crashes <- transform(crashes,
                  day.part = as.factor(mapvalues(HOUR_OF_DAY, c(0:23, 99),
                                                 c(rep("night", 5), rep("morning", 5), rep("mid", 6), rep("evening", 4), rep("night", 4), "unknown"))))

crashes <- transform(crashes, day.part = factor(day.part, levels = c("morning", "mid", "evening", "night", "unknown")))

# create commute variables
crashes <- transform(crashes,
                  commute = as.factor(mapvalues(HOUR_OF_DAY, c(0:23, 99),
                                                 c(rep("offpeak", 5), rep("commute", 5), rep("offpeak", 6), rep("commute", 4), rep("offpeak", 4), "unknown"))))


# label months
crashes <- transform(crashes,
                      CRASH_MONTH = mapvalues(CRASH_MONTH, c(1:12),
                                        c("Jan", "Feb", "Mar", "Apr", "May", "June",
                                          "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")))

# order months correctly
crashes <- transform(crashes, CRASH_MONTH = factor(CRASH_MONTH, levels = c("Jan", "Feb", "Mar", "Apr", "May",
                                                                 "June", "Jul", "Aug", "Sep", "Oct",
                                                                 "Nov", "Dec")))

# day of week labels
crashes <- transform(crashes, DAY_OF_WEEK = mapvalues(DAY_OF_WEEK, c(1:7),
                                                      c("Sun", "Mon", "Tue", "Wed", "Thr", "Fri", "Sat")))

# order days of the week correctly
crashes <- transform(crashes, DAY_OF_WEEK = factor(DAY_OF_WEEK, levels = c("Sun", "Mon", "Tue", "Wed", "Thr", "Fri", "Sat")))

# create season variables
crashes <- transform(crashes,
                      season = mapvalues(CRASH_MONTH, c("Jan", "Feb", "Mar", "Apr", "May",
                                                                 "June",
                                          "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"),
                                        c(rep("winter", 2), rep("spring", 3), rep("summer", 3), 
                                          rep("fall", 3), "winter")))

# create weekend variables
crashes <- transform(crashes,
                  day.type = mapvalues(DAY_OF_WEEK, c("Sun", "Mon", "Tue", "Wed", "Thr", "Fri", "Sat"),
                                        c("weekend", rep("weekday", 5), "weekend")))


# After cleaning and creating new variables, subset bike and ped crashes

bikesPGH <- subset(crashes, subset = BICYCLE==1)
bikesPGH <- select(bikesPGH, CRASH_CRN, DISTRICT, CRASH_YEAR, HOUR_OF_DAY, PERSON_COUNT,
                   VEHICLE_COUNT, 
                   BICYCLE_COUNT, FATAL_COUNT, MOD_INJ_COUNT, MIN_INJ_COUNT,
                   BICYCLE_DEATH_COUNT, 
                   BICYCLE_MAJ_INJ_COUNT, PED_COUNT, PED_DEATH_COUNT, PED_MAJ_INJ_COUNT,
                   LATITUDE, LONGITUDE, 
                   DEC_LAT, DEC_LONG, ALCOHOL_RELATED, DRINKING_DRIVER, CELL_PHONE,
                   CROSS_MEDIAN,
                   SPEEDING_RELATED,AGGRESSIVE_DRIVING, FATIGUE_ASLEEP, NHTSA_AGG_DRIVING, 
                   day.part, commute)

#bikes1115 <- filter(bikes, CRASH_YEAR %in% c(2011:2015))

#write.table(bikesPGH, "bikesPGHcrash.csv", row.names = F, col.names = T, sep=",")
ped <- subset(crashes, subset = COLLISION_TYPE==8)

# subset fatal or severe crashes
crashes.major.fatal <- subset(crashes, subset = FATAL_OR_MAJ_INJ=="1")
bikes.major <- subset(crashes.major.fatal, subset = BICYCLE==1)
ped.major <- subset(crashes.major.fatal, subset = COLLISION_TYPE==8)


# Population estimates from https://www.google.com/publicdata/explore?ds=kf7tgg1uo9ude_&ctype=l&strail=false&bcs=d&nselm=h&met_y=population&scale_y=lin&ind_y=false&rdim=county&idim=county:42003&ifdim=county:state:42000&hl=en&dl=en&ind=false&xMax=-73.14867066406248&xMin=-82.27830933593748&yMax=38.987950760624564&yMin=42.83844828764023&mapType=t&icfg&iconSize=0.5
years <- c(2004:2014)
county.pop <- c(1245900, 1232500, 1223200, 1219900, 1218200, 1218500, 1223800, 1227500, 1230400, 1233000, 1231300)
county.pop.years <- cbind(years, county.pop)
county.pop.years <- data.frame(county.pop.years)


# table of crash frequency per 1000 residents
# create freqency tables by year
crash.freq <- table(crashes$CRASH_YEAR)
crash.freq <- data.frame(crash.freq)
bike.freq <- table(bikes$CRASH_YEAR)
bike.freq <- data.frame(bike.freq)
ped.freq <- table(ped$CRASH_YEAR)
ped.freq <- data.frame(ped.freq)
  # create column names
colnames(crash.freq) <- c("years", "acc.count")
colnames(bike.freq) <- c("years", "bike.count")
colnames(ped.freq) <- c("years", "ped.count")
  # join data to population
crash.freq.pop <- join(crash.freq, county.pop.years, by="years")
crash.freq.pop <- join(bike.freq, crash.freq.pop, by="years")
crash.freq.pop <- join(ped.freq, crash.freq.pop, by="years")
  # create frequency per 1000 residents columns
crash.freq.pop$acc.per.1000 <- (crash.freq.pop$acc.count/crash.freq.pop$county.pop)*1000
crash.freq.pop$bikeacc.per.1000 <- (crash.freq.pop$bike.count/crash.freq.pop$county.pop)*1000
crash.freq.pop$pedacc.per.1000 <- (crash.freq.pop$ped.count/crash.freq.pop$county.pop)*1000
crash.freq.pop$caracc.per.bikeacc <- (crash.freq.pop$acc.count/crash.freq.pop$bike.count)
crash.freq.pop$caracc.per.pedacc <- (crash.freq.pop$acc.count/crash.freq.pop$ped.count)


```


### Project Introduction
This project analyzes 10 years of crash data from Allegheny County, over 132,000 incidents, retrieved from the [WPRDC](https://data.wprdc.org/dataset/allegheny-county-crash-data).

Crash data are information that derives from the information that comes from a reportable crash, generally meaning someone was injured or killed, or there was “damage to any vehicle to the extent that it cannot be driven under it’s own power in it’s customary manner without further damage or hazard to the vehicle”. Crash data does not include non-reportable crashesor near misses.

### Methodology
summary


### Preliminary Exploration of the Data

Overall:

  - Crashes with at least one minor injury make up `r round((table(crashes$MINOR_INJURY)[2])/nrow(crashes)*100, 1)`% of incidents, including `r round(100-(table(crashes$MOD_INJ_COUNT)[1])/nrow(crashes)*100, 1)`% with at least moderate injuries and only `r round(100-(table(crashes$BELTED_MAJ_INJ_COUNT)[1])/nrow(crashes)*100, 1)`% with major injuries. Crashes with fatalities make up `r round(100-(table(crashes$FATAL_COUNT)[1])/nrow(crashes)*100, 1)`% of incidents.
  - Drivers were drinking in `r round((table(crashes$DRINKING_DRIVER)[2])/nrow(crashes)*100, 1)`% of crashes.
  - `r round((table(crashes$DRIVER_16YR)[2])/nrow(crashes)*100, 1)`% of crashes involved 16 year old drivers, `r round((table(crashes$DRIVER_17YR)[2])/nrow(crashes)*100, 1)`% of crashes involved 17 year old drivers
  - `r round((table(crashes$DRIVER_65_74YR)[2])/nrow(crashes)*100, 1)`% of crashes involved drivers between ages 65 to 74 with `r round((table(crashes$DRIVER_75PLUS)[2])/nrow(crashes)*100, 1)`% of crashes with drivers aged 75 or older.
  
  
Let's break down the crash counts for the years we have data, 2004 to 2014, and compare to the county population.

#### Year to Year Comparisons

```{r population_graphs}
yearbreaks <- c(2004, 2006, 2008, 2010, 2012, 2014)

g1 <- qplot(crashes$CRASH_YEAR, 
      fill = I(cbPalette[3]),
      xlab = "Year",
      main = "All Crash Count by Year",
      ylab = "All Crashes",
      geom = "histogram",
      binwidth=1,
      col=I("white"),
      xlim=c(2002.5,2016.5)
      ) + 
  scale_x_continuous(breaks = yearbreaks)

pop.graph <- qplot(data=crash.freq.pop,
            y=county.pop,
            x=years,
            xlab = "Year",
            ylab = "Population",
            main = "County Population"
            ) + 
  scale_y_continuous(limits = c(1100000, 1250000)) + 
  scale_x_discrete(breaks = yearbreaks) +
  geom_bar(stat="identity", fill = I(cbPalette[8]))

grid.arrange(g1, pop.graph, ncol=2)
```

We see some variation in the number of crashes each year, but it does not appear to be correlated to the county's population. Let's break out bike and pedestrian crashes as well. Here we see a significant dip in vehicles hitting pedestrians in 2009, a dip in car-bike incidents in 2010 with some other peaks and valleys.

```{r pedbike_graphs_years}
g2 <- qplot(bikes$CRASH_YEAR, 
      fill = I(cbPalette[7]),
      xlab = "Year",
      ylab = "Bike Crashes",
      main = "Bike Crash Count by Year",
      geom = "histogram",
      binwidth=1,
      col=I("white")
      ) + scale_x_continuous(breaks = yearbreaks)

g3 <- qplot(ped$CRASH_YEAR, 
      fill = I(cbPalette[2]),
      xlab = "Year",
      ylab = "Pedestrian Crashes",
      main = "Ped Crash Count by Year",
      geom = "histogram",
      binwidth=1,
      col=I("white")
      ) + scale_x_continuous(breaks = yearbreaks)

grid.arrange(g2, g3, ncol=2)
```

Let's look at crash frequency per 1000 residents (note differing scales for each type).

**Add in layered histogram? All three on one graph?**

```{r frequency_graphs}
g4 <- qplot(data=crash.freq.pop,
            y=acc.per.1000,
            x=years,
            xlab = "Year",
            ylab = "Crashes per 1000 Residents",
            main = "All Crash Frequency"
            ) +
  scale_y_continuous(limits = c(0, 11)) +
  scale_x_discrete(breaks = yearbreaks) +
  geom_bar(stat="identity", fill = I(cbPalette[3]))


g5 <- qplot(data=crash.freq.pop,
            y=bikeacc.per.1000,
            x=years,
            xlab = "Year",
            ylab = "Crashes per 1000 Residents",
            main = "Car-Bike Frequency"
            ) +
  scale_y_continuous(limits = c(0, 0.125)) +
  scale_x_discrete(breaks = yearbreaks) +
  geom_bar(stat="identity", fill = I(cbPalette[7]))

g6 <- qplot(data=crash.freq.pop,
            y=pedacc.per.1000,
            x=years,
            xlab = "Year",
            ylab = "Crashes per 1000 Residents",
            main = "Car-Ped Frequency"
            ) +
  scale_y_continuous(limits = c(0, 0.45)) +
  scale_x_discrete(breaks = yearbreaks) +
  geom_bar(stat="identity", fill = I(cbPalette[2]))

grid.arrange(g4, g5, g6, ncol=3)
```

Let's also look at the differences between "Fatality or Major Injury" crashes and other crashes, which make up `r round((table(crashes$FATAL_OR_MAJ_INJ)[2])/nrow(crashes)*100, 2)`% of incidents.

```{r severe_days_of_week}

b1 <- qplot(crashes$DAY_OF_WEEK, 
      fill = I(cbPalette[4]),
      xlab = "Day",
      main = "All Reported")

b2 <- qplot(crashes.major.fatal$DAY_OF_WEEK, 
      fill = I(cbPalette[6]),
      xlab = "Day",
      main = "Severe & Fatal Only")

b3 <- qplot(crashes$day.type, 
      fill = I(cbPalette[4]),
      main = "All Reported",
      xlab = "Day Type")
b4 <- qplot(crashes.major.fatal$day.type, 
      fill = I(cbPalette[6]),
      main = "Severe & Fatal Only",
      xlab = "Day Type")

grid.arrange(b1, b2, b3, b4, ncol=2)

b5 <- qplot(crashes$day.part, 
      fill = I(cbPalette[4]),
      xlab = "Time of Day",
      main = "All Reported")
b6 <- qplot(crashes.major.fatal$day.part, 
      fill = I(cbPalette[6]),
      xlab = "Time of Day",
      main = "Fatalities Only")

b7 <- qplot(crashes$commute, 
      fill = I(cbPalette[4]),
      main = "All Reported",
      xlab = "Commute")
b8 <- qplot(crashes.major.fatal$commute, 
      fill = I(cbPalette[6]),
      main = "Fatalities Only",
      xlab = "Commute")

grid.arrange(b5, b6, b7, b8, ncol=2)
```


#### Month to Month and Seasonal Comparisons

Breaking out by months, there appear to be more crashes in January and December than the rest of year, which might be because of winter weather conditions on the road.

If we again break out crashes involving bicyclist and pedestrians, we see a clear trend that there more crashes involving bicycles in warmer months. This makes sense -- there are probably more riders. We don't necessarily need to normalized this by vehicle miles traveled, though, it tells us what time of year to anticipate more incidents.

```{r pedbike_graphs_months}
monthbreaks = c("Jan", "Mar", "May", "Jul", "Sep", "Nov")

s1 <- qplot(crashes$CRASH_MONTH, 
      fill = I(cbPalette[3]),
      xlab = "Month",
      main = "All Reported") + 
  scale_x_discrete(breaks = monthbreaks)

s2 <- qplot(bikes$CRASH_MONTH, 
      fill = I(cbPalette[7]),
      xlab = "Month",
      main = "Car-Bike by Month") + 
  scale_x_discrete(breaks = monthbreaks)

s3 <- qplot(ped$CRASH_MONTH, 
      fill = I(cbPalette[2]),
      xlab = "Month",
      main = "Car-Ped by Month") + 
  scale_x_discrete(breaks = monthbreaks)

grid.arrange(s1, s2, s3, ncol=3)
```

Again let's also look at the differences between "Fatality or Major Injury" crashes and other crashes.

```{r severe_month}

s4 <- qplot(crashes$CRASH_MONTH, 
      fill = I(cbPalette[4]),
      xlab = "Month",
      main = "All Reported") + 
  scale_x_discrete(breaks = monthbreaks)

s5 <- qplot(crashes.major.fatal$CRASH_MONTH, 
      fill = I(cbPalette[6]),
      xlab = "Month",
      main = "Severe & Fatal Only") + 
  scale_x_discrete(breaks = monthbreaks)

#grid.arrange(s4, s5, ncol=2)

s6 <- qplot(bikes$CRASH_MONTH, 
      fill = I(cbPalette[4]),
      xlab = "Month",
      main = "All Car-Bike") + 
  scale_x_discrete(breaks = monthbreaks)

s7 <- qplot(bikes.major$CRASH_MONTH, 
      fill = I(cbPalette[6]),
      xlab = "Month",
      main = "Car-Bike Severe Fatal") + 
  scale_x_discrete(breaks = monthbreaks)

s8 <- qplot(ped$CRASH_MONTH, 
      fill = I(cbPalette[4]),
      xlab = "Month",
      main = "All Car-Ped") + 
  scale_x_discrete(breaks = monthbreaks)

s9 <- qplot(ped.major$CRASH_MONTH, 
      fill = I(cbPalette[6]),
      xlab = "Month",
      main = "Car-Ped Severe Fatal") + 
  scale_x_discrete(breaks = monthbreaks)

grid.arrange(s4, s5, s6, s7, s8, s9, ncol=2)
```




```{r cellphones}
# table(crashes$CELL_PHONE, crashes$FATAL_OR_MAJ_INJ)
table(crashes$DRIVER_17YR)

```


```{r pressure}
# g2 <- qplot(road.fatal$day, 
#       fill = I(cbPalette[3]),
#       xlab = "Day",
#       main = "Fatalities Only")
# 
# g3 <- qplot(road$day.type, 
#       fill = I(cbPalette[7]),
#       main = "All Reported")
# g4 <- qplot(road.fatal$day.type, 
#       fill = I(cbPalette[3]),
#       main = "Fatalities Only")

# grid.arrange(g1, g2, g3, g4, ncol=2)
```

